`respR` - An R package for the analysis of respirometry data

### 2017-11-01 manuscript/manuscript-HTML.Rmd
manuscript update

Depending on the experiment, different metabolic rate parameters may be of interest to researchers. These include maximum metabolic rates (MMR, or $MO_{2,max}$), rates under high activity or exhaustive exercise, and minimal metabolic rates ($MO_{2,min}$), which, depending on the organism, may be termed standard (SMR), basal (BMR), or resting metabolic rates, and typically represent the minimum metabolic cost of maintaining biological functioning (White and Kearney, 2014). Routine metabolic rate (RMR), is similar to $MO_{2,min}$, but takes into account that in some organisms energy is expended on small, sponateous movements to maintain posture or position (Rogers et al. 2016). Another metric of great interest is the critical oxygen tension, which represents the lowest level of oxygen at which aerobic metabolism is independent of the ambient partial pressure of oxygen ($P_{crit}$; Yeager & Ultsch, 1989, Hochachka and Somero, 2002).

@@

Recently, a number of open-source R software packages have become available which are designed for, or are suitable to analyse respirometry data. The `respirometry` package (https://CRAN.R-project.org/package=respirometry) contains a comprehensive collection of tools to explore and evaluate experimental parameters in aquatic respirometry and is useful for the design and diagnosis of experimental setups. The `rMR` package (https://CRAN.R-project.org/package=rMR) performs interval-based metabolic rate calculations and has a method to automatically detect $P_{crit}$ using the "broken-stick" regression method (Yeager & Ultsch, 1989). `LoLinR` (Olito et al., 2017), while not strictly coded with respirometry in mind, provides a statistically robust method of detecting a "best-fit" regression that applies to the calculation of metabolic rate, and performs very well at identifying truly linear subsets of a data series. Importantly, the packages above all assume that data are monotonic with evenly-spaced sampling periods with respect to time and do not have the tools to analyse complicated datasets that require extraction of multiple respirometry parameters.

Here we describe the R package `respR`, a set of functions desgined to provide an efficient, structured, and reproduceable workflow for the analysis of respirometry-related data (Fig. 1). The package contains utilities to: (1) analyse closed, intermittent, flow-through and open respirometry data, (2) determine volume, mass, and surface area specific oxygen uptake rates, (3) automatically detect maximal and minimal rates, $P_{crit}$, or the most linear results in large datasets with the help of traditional rolling regression and kernel density estimation techniques. Other functions useful in processing respirometry data are avaliable, and are described in more detail in our online html vignette (https://januarharianto.github.io/respR/). The package has a strong focus on aquatic respirometry, however similar principals apply to aerial respirometry, so the analysis workflow should also be suitable for these data. We demonstrate the utility of the package by analysing three datasets as case studies and proof of concept. These data were collected using closed, intermittent and flow-through respirometry, and we show how `respR` might be used to determine $MO_{2,max}$, $MO_{2,min}$, and $P_{crit}$.

@@

Data should be formatted correctly before use in `respR`. The function `check.input()` performs error checks on a data frame of any size, and can extract a two-column data frame for data exploration and subsequent analyses. Data must be numeric, because `respR` has limited support for date/time data (e.g. `POSIXct` and `POSIXlt` classes) and performing calculations on different time formats have many caveats that are beyond the scope of the package. Time data, being a continuous variable, should be sequential, without duplicates. Ideally, time data should also be evenly-spaced, that is, the time period between samples are equal, as this will affect data-extraction methods later on if data is analysed by sample length instead of time. Error check results are summarised in the R console (see supplementary material) and plot of the data is shown for visual inspection (e.g. Fig. 1).

It should be noted that invoking `check.input` is optional -- the main functions in our package will readily accept any data frame as long as data are numeric. Running `check.input` is an exploratory step that flags potential issues about the data before it is further analysed.

@@

Background $RO_2$ is often measured to account for the influence of bacterial respiration in the chamber, and applied as a correction to the main experiment (Rodgers et al., 2016). Since background rates typically account for a small percentage of experimental rates, multiple “blank” experiments are routinely conducted and the rates are averaged across several datasets to obtain a more accurate estimate of the correction (e.g. Carey and Sigwart, 2014; Daoud et al., 2007). The function `calc.bg.rate()` uses simple linear regression to simultaneously process multiple background rate measurements and produces an output object that can be used in later functions that accept background rate arguments. Because the oxygen consumption units are non-volume and non-mass specific at the point of correction, the functions in `respR` can adjust $RO_2$ directly without bias: $$RO_{2(correction)} = RO_2 - RO_{2(bg)}$$

### Extracting data for $RO_2$

Typically, the analysis respirometry data requires the selection one or more linear sections of the raw data for more accurate calculations of $RO_2$ (e.g. Gordon et al., 1989, Dorey et al., 2014, Chu & Gal, 2017). The function `calc.rate()` can manually extract and process data segments from respiromety data. Data segments can be selected by (1) time period, (2) row numbers, (3) $O_2$ decrease or (4) proportion, which should accomodate most, if not all, data selection requirements and allow for consistent reporting of methods and results (e.g. estimating $RO_2$ over a regular 10-minute time window across multiple specimens). Rates are determined using linear regression, however a two-point analysis is also included alongside all regression results.

For more complex respirometry parameters, the function `auto.rate()` uses rolling regression and kernel density estimate algorithms to detect patterns in respirometry data. It can calculate maximum or minimum rates, extract rates at non-overlapping intervals, and has an exploratory method to detect the most linear sections of the data. When called, the function takes the data frame of length $m$, and performs a rolling regression of sample size $n$ iteratively across the length of the data (Fig. 2b). Thus, a total of $(m−n)+1$ number of overlapping regressions are fitted, and we obtain a new dataset of regressions as a function of time (rolling regression, Fig. 3b), which is then ranked by size to obtain maximum and minimum values.  There are clear advantages to using `auto.rate()` when estimating parameters for $MO_{2,max}$ or $MO_{2,min}$. Unlike manual methods, where data may be arbitrary selected, `auto.rate()` adheres to well-defined rules to obtain the rate of interest, so its methods are results are transparent and fully reproducible. To estimate the most linear sections of the data, further computations are performed after rolling regressions. Logically, linear sections of the data should reflect a consistently stable (i.e. flat) rate. The function takes advantage of this relationship and performs a Gaussian kernel density estimate of all regressions (see Silverman, 1986). New datasets around the kernel bandwidth are then resampled around peak values of the generated model, re-analysed, and then ranked by size to determine the most linear sections.

The function `calc.rate.ft()` is used to specifically analyse flowthrough respirometry data, and is similar to `calc.rate()` in functionality. However it accepts additional inputs of incurrent and excurrent oxygen concentration, and flow rate, that are not available in `calc.rate()`. Results in `calc.rate.ft()` are processed to determine the mean or modal $RO_2$.

@@

We provide two methods of estimating $P_{crit}$. The first is a "broken-stick" regression (BSR) approach, adopted from Yeager and Ultsch (1989) in which two segments of the data are iteratively fitted and the intersection with the smallest sum of the residual sum of squares between the two linear models is the estimated critical point. The second method is the segmented, or "broken-line" regression approach, presented by Muggeo (2003) which estimates the critical point by iteratively fitting two intersecting models on the data, and picking the point that minimises the "gap" between the two fitted lines. To determine $P_{crit}$, $RO_2$ (or $MO_2$) and ambient oxygen concentration ($PO_2$) must be provided. If $RO_2$ has not been calculated, `pcrit()` can automatically perform a rolling regression directly from raw respirometry data to generate the data and arrange it against $PO_2$, and then use one of the two methods to estimate the value (Fig. 4).

@@

For delibrate reasons, units of measurements are not used in any of the functions `calc.bg.rate()`, `calc.rate()`, `auto.rate()`, `calc.rate.ft()` and `pcrit()`. Instead, two functions, `convert.do()` and `calc.mo2()`, need to be called separately to convert units and to perform volumetric or mass-specific scaling. This unique approach has the advantage of allowing us to perform "dimensionless" rate calculations for data that may not be fully supported by the entire `respR` workflow. For example, measurements of respiration can be scaled to surface area (diatoms, Iversen & Ploug, 2013, Naumann et al., 2014), volume (copepod eggs, Hammervold et al., 2015), or density (sea urchin larvae, Stumpp et al., 2011).

The function `convert.do()` is used to convert units of oxygen concentration (e.g. mg/L or %), while `calc.mo2()` directly accepts object outputs from prior rate-calculating functions and performs unit conversions, volumetric and mass-specific conversions at the same time.

@@

# Concluding remarks and future improvements

@@

```{r, echo = F, out.width = "900px"}
knitr::include_graphics("figs/fig1.png")
```

@@

```{r, echo = F, out.width = "650px"}
knitr::include_graphics("figs/fig2.png")
```

**Fig. 2** Simplified illustration showing the sampling windows used by `auto.rate()` during **(a)** interval regression analysis, where sample windows do not overlap and **(b)** rolling regression analysis, where sample windows overlap and move forward by one sample unit at a time.

```{r, echo = F, , out.width = "800px"}
knitr::include_graphics("figs/fig3.png")
```

**Fig. 3** Selected output plots from the function `calc.rate()` for the analysis of the data frame `urchin2013`.

>>>>

Depending on the experiment, different metabolic rate parameters may be of interest to researchers. These include maximum metabolic rates (MMR, or $MO_{2,max}$), which define rates under high activity or exhaustive exercise, and minimal metabolic rates ($MO_{2,min}$), which may be termed standard (SMR), basal (BMR), or resting metabolic rates, and typically represent the minimum metabolic cost of maintaining biological functioning (White and Kearney, 2014). Routine metabolic rate (RMR), is similar to $MO_{2,min}$, but takes into account that in some organisms energy is expended on small, sponateous movements to maintain posture or position (Rogers et al. 2016). Another metric of great interest is the critical oxygen tension, which represents the lowest level of oxygen at which aerobic metabolism is independent of the ambient partial pressure of oxygen ($P_{crit}$; Yeager & Ultsch, 1989, Hochachka and Somero, 2002).

@@

Recently, a number of open-source R software packages have become available which are designed for, or are suitable to analyse respirometry data. The `respirometry` package (https://CRAN.R-project.org/package=respirometry) contains a comprehensive collection of tools to explore and evaluate experimental parameters in aquatic respirometry and is useful for the design and diagnosis of experimental setups. The `rMR` package (https://CRAN.R-project.org/package=rMR) performs interval-based metabolic rate calculations and has a method to automatically detect $P_{crit}$ using the "broken-stick" regression method (Yeager & Ultsch, 1989). `LoLinR` (Olito et al., 2017), while not strictly coded for respirometry, provides a statistically robust method of detecting a "best-fit" regression that applies to the calculation of metabolic rate, and performs very well at identifying truly linear subsets of a data series. Importantly, all of these packages assume that data are monotonic with evenly-spaced sampling periods with respect to time, and do not have the tools to analyse complicated datasets to automatically determine useful metric such as $MO_{2,max}$ and $MO_{2,min}$.

<!-- FIGURE 1 -->
Here we describe the R package `respR`, a set of functions desgined to provide an efficient, structured, and reproduceable workflow for the analysis of respirometry-related data (**Table 1**). The package contains utilities to: (1) analyse closed, intermittent, flow-through and open respirometry data, (2) determine volume, mass, and surface area specific oxygen uptake rates, (3) automatically detect maximal and minimal rates, $P_{crit}$, or (4) detect the most linear sections of large datasets, with the help of traditional rolling regression and kernel density estimation techniques. Other functions useful in processing respirometry data are avaliable and are described in more detail in our online html vignette (https://januarharianto.github.io/respR/). The package has a strong focus on aquatic respirometry, however similar principals apply to aerial respirometry, so the analysis workflow should also be suitable for these data. We demonstrate the utility of the package by analysing real data that were collected using closed, intermittent and flow-through respirometry as proof of concept, and show how `respR` might be used to determine $MO_{2,max}$, $MO_{2,min}$, and $P_{crit}$.

@@

Data should be formatted correctly before use in `respR`. The function `inspect.data()` performs error checks on a data frame of any size, and can extract a two-column data frame for data exploration and subsequent analyses. Ideally all data are numeric, because `respR` has limited support for date/time data (e.g. `POSIXct` and `POSIXlt` classes), and performing calculations on different time formats have many caveats that are beyond the scope of the package. Time data, being a continuous variable, should be sequential, without duplicates. Ideally, sampling frequency should be monotonic, but in some cases this may be difficult as machine drift and/or data dropouts may occur. However, irregular time series data can still be analysed by `respR` as it supports time-based data extraction methods.

It should be noted that using `inspect.data()` is optional -- the main functions in our package will readily work with data frames that were not checked. Running `inspect.data()` is an exploratory step that flags potential issues about the data before they are analysed. Error check results are summarised in the R console, and all error locations (i.e. row numbers) are saved in the output for inspection. A plot of the data is also shown for visual inspection (**Fig. 1**).

@@

Background $RO_2$ is often measured to account for the influence of bacterial respiration in the chamber, and applied as a correction to the main experiment (Rodgers et al., 2016). Since background rates typically account for a small percentage of experimental rates, multiple “blank” experiments are routinely conducted and the rates are averaged across several datasets to obtain a more accurate estimate of the correction (e.g. Carey and Sigwart, 2014; Daoud et al., 2007). The function `calc.rate.bg()` uses simple linear regression to simultaneously process multiple background rate measurements and produces an output object that can be used directly in later functions that accept background rate arguments.  Data segments can be truncated before analysis, by either time period or row numbers. As the oxygen consumption units are not mass-specific at the point of correction, the functions in `respR` can adjust $RO_2$ directly without bias using the equation: $$RO_{2(correction)} = RO_2 - RO_{2(bg)}$$

### Extracting data and calculating $RO_2$

Typically, the analysis respirometry data requires the selection one or more linear sections of the raw data for more accurate calculations of $RO_2$ (e.g. Gordon et al., 1989, Dorey et al., 2014, Chu & Gal, 2017). The function `calc.rate()` can manually extract and process data segments from respiromety data. Data segments can be selected by (1) time period, (2) row numbers, (3) $O_2$ decrease or (4) proportion, which should accomodate most, if not all, data selection requirements and allow for consistent reporting of methods and results (e.g. estimating $RO_2$ over a fixed 10-minute time window across multiple specimens). Rates are determined using linear regression, however a two-point analysis is also included alongside all regression results.

For more complex respirometry parameters, the function `auto.rate()` uses rolling regression and kernel density estimate algorithms to detect patterns in respirometry data. It can calculate maximum or minimum rates, extract rates at non-overlapping intervals, and has an exploratory method to detect the most linear sections of the data. When called, the function takes the data frame of length $m$, and performs a rolling regression of sample size $n$ iteratively across the length of the data (Fig. 2a). Thus, a total of $(m−n)+1$ number of overlapping regressions are fitted, and we obtain a new dataset of regressions as a function of time (rolling regression, Fig. 3b), which is then ranked by size to obtain maximum and minimum values. If an interval-based regression is required, the function selects non-overlapping sections of the data for regressions (Fig. 2b).

To estimate the most linear sections of the data, further computations are performed after rolling regressions. Logically, linear sections of the data should reflect a consistently stable (i.e. flat) rate. If the rate changes at some point during the sampling period, thent the rolling estimate should detect the variability. The function takes advantage of this relationship and performs a Gaussian kernel density estimate (KDE) of all regressions (see Silverman, 1986). Peaks in the density thus reflect stable rates, and new segments of the data within the kernel bandwidth of the peak values are resampled and re-analysed to determine the most linear sections.

There are clear advantages to using `auto.rate()` when estimating parameters for $MO_{2,max}$ or $MO_{2,min}$. Unlike manual methods, where data may be arbitrary selected, `auto.rate()` adheres to well-defined rules to obtain the rate of interest, so its methods are results are transparent and fully reproducible.

The function `calc.rate.ft()` is used to specifically analyse flowthrough respirometry data, and is similar to `calc.rate()` in functionality. However it accepts additional inputs of incurrent and excurrent oxygen concentration, and flow rate, that are not available in `calc.rate()`.`calc.rate.ft()` can process data to determine the mean or modal $RO_2$ over a period.

@@

We provide two methods of estimating $P_{crit}$. The first is a "broken-stick" regression (BSR) approach, adopted from Yeager and Ultsch (1989) in which two segments of the data are iteratively fitted and the intersection with the smallest sum of the residual sum of squares between the two linear models is the estimated critical point. The second method is the segmented, or "broken-line" regression approach, presented by Muggeo (2003) which estimates the critical point by iteratively fitting two intersecting models on the data, and picking the point that minimises the "gap" between the two fitted lines.

To determine $P_{crit}$, $RO_2$ (or $MO_2$) and ambient oxygen concentration ($PO_2$) must be provided. If $RO_2$ is not available, but the respirometry data exists to calculate it, `pcrit()` can automatically perform a rolling regression directly on the data to generate $RO_2$ and arrange it against the rolling mean of $PO_2$. This method is unique to `respR` -- in the past, a user may generate the same by cutting the data into equal sections and performing a linear regression on every section. A rolling regression

@@

For delibrate reasons, units of measurements are not used in any of the functions `calc.rate.bg()`, `calc.rate()`, `calc.rate.ft()`, `auto.rate()` and `pcrit()`. Instead, two functions, `convert.rate()` and `scale.rate()`, need to be called separately to convert units and to perform volumetric or mass-specific scaling. This unique approach has the advantage of allowing us to perform "dimensionless" rate calculations for data that may not be fully supported by the entire `respR` workflow. For example, measurements of respiration can be scaled to surface area (diatoms, Iversen & Ploug, 2013, Naumann et al., 2014), volume (copepod eggs, Hammervold et al., 2015), or density (sea urchin larvae, Stumpp et al., 2011). Users may thus first perform rate calculations using `respR`, and then scale the results manually after if there is no support for their specific scaling method.

The function `convert.rate()` is used to convert units of oxygen concentration (e.g. mg/L or %) only and has support for over 132 unit combinations. As unit conventions differ between research disciplines and we do not expect

`scale.rate()` directly accepts object outputs from prior rate-calculating functions and performs volumetric and mass-specific conversions. We have implemented a "semi-fuzzy" approach to

@@

# Concluding remarks and future improvements
While performing the required linear regressions to estimate the parameter of interest is a relatively simple task, a considerable amount of time is spent in data formatting and manipulation.

@@

```{r, echo = F, out.width = "900px"}
knitr::include_graphics("figs/Figure 1.png")
```

@@

```{r, echo = F, out.width = "650px"}
knitr::include_graphics("figs/Figure 2.png")
```

**Fig. 2** Illustration showing the sampling method used by `auto.rate()` during **(a)** interval regression analysis, where sample windows do not overlap and **(b)** rolling regression analysis, where sample windows overlap and move forward by one sample unit at a time.

```{r, echo = F, , out.width = "800px"}
knitr::include_graphics("figs/Figure 3.png")
```

**Fig. 3** Typical graphical outputs of the function `auto.rate()`. (a) Depending on the `method` input, the function is able to show: (clockwise) the maximum rate detected, the minimum rate detected, calculation of rate by fixed-width intervals, and the most linear region as determined by kernel density estimates. (b) In addition, the outputs will also print diagnostic plots of: (clockwise) the rolling regression of rate as a function of time with dotted lines indicating the region used to estimate the parameter, a kernel density plot that shows only when the `linear` method is called, a residual plot of the regression, and a normal Q-Q plot of the residuals of the regression. All data is obtained from the example dataset `sardine` which is included in the package.
